Recursive partitioning of linear growth curve models

### 2022-11-11 pkg/glmertree/paper growth curves/_Partitioning_GCMs_with_GLMM_trees.Rnw
improvements in abstract and intro
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3231 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: ea3f202bd8f6cc54295e8a79410a244426cf5b26

\title{Recursive partitioning of linear growth curve models}
\shorttitle{Partitioning linear trajectories with GLMM trees}
\authorsnames{Marjolein Fokkema$^1$, Achim Zeileis$^2$}
\authorsaffiliations{{$^1$Leiden University}, {$^2$University of Innbruck}}

\abstract{Growth curve models are popular tools for studying development over time. Often, researchers have a specific interest in uncovering subgroups that show different patterns of growth over time. Recursive partitioning methods (RPMs) allow for detecting such subgroups. In this paper, we focus on the use of generalized linear mixed effects model trees (GLMM trees) for partitioning LGCMs. GLMM trees were originally proposed for subgroup detection in clustered cross-sectional data. Here, we introduce recent adaptations of the algorithm that that are of particular relevance for partitioning longitudinal models, and growth curve models in particulat. Using simulated and existing data, we compare the performance of GLMM trees and other methods for partitioning LGCMs: SEM trees, LongCART and longRPart. We find that GLMM trees provide similar accuracy compared to SEM trees, and better accuracy compared to LongCART and longRPart. In addition, GLMM trees allow for the analysis of trajectories with both discrete and continuous time, are less sensitive to (mis-)specification of the random-effects structure and require substantially lower computation times.}

@@

In longitudinal studies, (generalized) linear growth curve modeling is often used to model change over time, and inter individual differences in change. For example, researchers may want to model student's reading or math development in an educational study, or symptom reduction over the course of a clinical trail. Inter individual heterogeneity is likely in such studies, where (groups of) participants show higher or lower rates of growth over time. Through the use of mixed-effects or latent-variable models, such heterogeneity can be captured, and explained by covariates of a-priori known relevance \citep{NeisyMatt18}.

Often, the covariates that predict or affect the trajectories may not be known a-priori, or researchers may have a specific interest in finding subgroups of participants with similar trajectories. Recursive partitioning methods (RPMs), also known as decision-tree methods, are particularly suited in such cases. Several existing RPMs allow for partitioning growth trajectories: For example, GUIDE \citep{Loh02}, longRPart \citep{AbdoyLeBl02}, GEE-based decision trees \citep{Lee05}, longitudinal interaction trees \cite[IT; ][]{SuyMene11}, Structural Equation Model (SEM) trees \citep{BranyOert13, ArnoyVoel21}, mixed-effects longitudinal trees \citep[MELT; ][]{EoyCho14} and LongCART \cite{KundyHare19} allow for partitioning linear growth curve models (LGCMs). Partitioning based on non-linear growth curve models can be performed with the longRPart2 \citep{StegyJaco18} and IT-LT \citep{WeiyLiu20} methods\footnote{Both IT methods specifically focus at detecting so-called \textit{treatment-subgroup interactions}: subgroups that respond more favorably to one treatment than another.}.

In this paper, we focus on generalized linear mixed-effects model trees \citep[GLMM trees; ][]{FokkySmit18}, which allow for subgroup detection in a wide range of mixed-effects models. Although originally not specifically developed for partitioning LGCMs, application of GLMM trees for partitioning LGCMs is straightforward. We propose two adaptations to the algorithm, which we expect to be of particular relevance for partitioning LGCMs.

The main difference between GLMM trees and the aforementioned RPMs, is that the latter all fit the full parametric model in each of the nodes. In contrast, GLMM trees uses the observations in each node to estimate the fixed-effects parameters, while using all observations to estimate the random-effects parameters. This may not only reduce computation time; it also reduces the total number of parameters to be estimated, which may improve stability and reduce overfitting. This local/global estimation approach was originally proposed by \cite{HajjyBell11} and \cite{SelaySimo12} for trees with constant fits in the terminal nodes; \cite{FokkySmit18} generalized it to model-based trees.

@@

LongCART, MELT, GEE-based decision trees and score-based SEM trees fit full parametric models in each of the nodes as well, but do not refit models for cutpoint selection; they employ the predictions or residuals from the fitted model in the current node for selecting the best split. This reduces computational burden. Furthermore, these approaches allow for employing a two-step approach for split detection, which separates variable and cutpoint selection, thus preventing variable selection bias. The GLMM tree algorithm is more similar to the latter group of RPMs, in that it employs a two-step approach to split detection.

The remainder of this paper is structured as follows: In the next section (Estimation) we explain how GLMM trees are estimated, and how estimation can be adjusted to (better) account for dependence structures encountered in LGCMs. Next, in Study I, we perform a simulation study to assess performance of GLMM trees and the proposed adaptations in partitioning LGCMs. In Study II, we use the same simulation design to compare the performance of GLMM trees with that of two other mixed-effects RPMs: SEM trees and LongCART.  In Study III, we compare the performance of GLMM trees and longRPart in partitioning children's trajectories of reading, math and science abilities. In the Discussion, we summarize our findings and discuss implications.

>>>>

\title{Recursive partitioning of linear growth curve models}
\shorttitle{Recursive partitioning of linear growth curve models}
\authorsnames{Marjolein Fokkema$^1$, Achim Zeileis$^2$}
\authorsaffiliations{{$^1$Leiden University}, {$^2$University of Innbruck}}

\abstract{Growth curve models are popular tools for studying {\color{red!80} the development of a response variable within the same observational units} over time. Often, researchers have a specific interest in uncovering subgroups that show different patterns of growth over time. Recursive partitioning methods (RPMs) allow for detecting such subgroups. In this paper, we focus on the use of generalized linear mixed effects model trees (GLMM trees) for partitioning linear growth curve models (LGCMs). GLMM trees were originally proposed for subgroup detection in clustered cross-sectional data. {\color{red!80} Here, we introduce an extension of the algorithm for longitudinal (as opposed to clustered) data that can detect subgroups defined by subject-specific covariates on/or time-varying covariates. The resulting extended GLMM trees are directly applicable to growth curve models as an important special case.} Using simulated and empirical data, we compare the performance of GLMM trees and other methods for partitioning LGCMs: SEM trees, LongCART and longRPart. We find that GLMM trees provide similar accuracy compared to SEM trees, and better accuracy compared to LongCART and longRPart. In addition, GLMM trees allow for the analysis of trajectories with both discrete and continuous time, are less sensitive to (mis-)specification of the random-effects structure and require substantially lower computation times.}

@@

In longitudinal studies, (generalized) linear growth curve modeling is often used to model change over time, and inter-individual differences in change. For example, researchers may want to model student's reading or math development in an educational study, or symptom reduction over the course of a clinical trail. Inter-individual heterogeneity is likely in such studies, where (groups of) participants show higher or lower rates of growth over time. Through the use of mixed-effects or latent-variable models, such heterogeneity can be captured, and explained by covariates of a-priori known relevance \citep{NeisyMatt18}.

Often, the covariates that predict or affect the trajectories may not be known a-priori, or researchers may have a specific interest in finding subgroups of participants with similar trajectories. Recursive partitioning methods (RPMs), also known as decision-tree methods, are particularly suited in such cases. Several existing RPMs allow for partitioning growth trajectories: For example, GUIDE \citep{Loh02}, longRPart \citep{AbdoyLeBl02}, GEE-based decision trees \citep{Lee05}, longitudinal interaction trees \cite[IT; ][]{SuyMene11}, structural equation model (SEM) trees \citep{BranyOert13, ArnoyVoel21}, mixed-effects longitudinal trees \citep[MELT; ][]{EoyCho14} and LongCART \cite{KundyHare19} allow for partitioning linear growth curve models (LGCMs). Partitioning based on non-linear growth curve models can be performed with the longRPart2 \citep{StegyJaco18} and IT-LT \citep{WeiyLiu20} methods\footnote{Both IT methods specifically focus at detecting so-called \textit{treatment-subgroup interactions}: subgroups that respond more favorably to one treatment than another.}.

In this paper, we focus on generalized linear mixed-effects model trees \citep[GLMM trees; ][]{FokkySmit18}, which allow for subgroup detection in a wide range of mixed-effects models.  {\color{red!80} FIXME: The rest of this paragraph needs to be a bit more forceful!} Although originally not specifically developed for partitioning LGCMs, application of GLMM trees for partitioning LGCMs is straightforward. We propose two adaptations to the algorithm, which we expect to be of particular relevance for partitioning LGCMs.

The main difference between GLMM trees and the aforementioned RPMs, is that the latter all fit the full parametric model in each of the nodes. In contrast, GLMM trees uses the observations in each node to estimate the fixed-effects parameters, while using all observations to estimate the random-effects parameters. This may not only reduce computation time; it also reduces the total number of parameters to be estimated, which may improve stability and reduce overfitting. This local/global estimation approach was originally proposed by \cite{HajjyBell11} and \cite{SelaySimo12} for trees with constant means in the terminal nodes; \cite{FokkySmit18} generalized it to trees containing generalized linear models in the nodes, i.e., potentially including regressors and/or using a non-Gaussian family.

@@

LongCART, MELT, GEE-based decision trees and score-based SEM trees fit full parametric models in each of the nodes as well, but do not refit models for cutpoint selection; they employ the predictions or residuals from the fitted model in the current node for selecting the best split. This reduces the computational burden. Furthermore, these approaches allow for employing a two-step approach for split detection, which separates variable and cutpoint selection, thus preventing variable selection bias. The GLMM tree algorithm is more similar to the latter group of RPMs, in that it employs a two-step approach to split detection.

The remainder of this paper is structured as follows: In the next section (Estimation) we explain how GLMM trees are estimated, and how estimation can be adjusted to (better) account for dependence structures encountered in LGCMs. Next, in Study I, we perform a simulation study to assess performance of GLMM trees and the proposed adaptations in partitioning LGCMs. In Study II, we use the same simulation design to compare the performance of GLMM trees with that of two other mixed-effects RPMs: SEM trees and LongCART.  In Study III, we compare the performance of GLMM trees and longRPart in partitioning children's trajectories of reading, math and science abilities. In the Discussion, we summarize our findings and their implications.
